---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
This is the report produced for the Peer Assignment 1 of "Reproducible Research" in Coursera.


## Loading and preprocessing the data

We load the data and look at the first 5 rows.
```{r}
setwd("/Users/isa/Documents/Courses/Johns Hopkins/Reproducible Research/Week2/RepData_PeerAssessment1")
data <- read.csv("activity.csv", colClasses = c("numeric", "factor", "numeric"))
head(data)
```

We transform the data to adequate format.
````{r}
library(lubridate)
data$date2=as.Date(strptime(data$date, format="%Y-%m-%d" ),"%Y%m%d")
str(data)
summary(data)
```


## What is mean total number of steps taken per day?
First, we aggregate the total number of steps taken per day and create a histogram that shows it.
````{r scatterplot, fig.width=12}
agg <- with(data, aggregate(data$steps, by=list(data$date2), sum))
barplot(agg$x, names.arg = agg$Group.1, col="red", main="Total number of steps taken per day", xlab="Days: From October 1st to December 1st",ylab="Steps Frequency")
```
We then calculate the mean and the median as shown respectively:
````{r}
mean(agg[, "x"], na.rm = TRUE)
median(agg[, "x"], na.rm = TRUE)
```


## What is the average daily activity pattern?
First, we aggregate the number of steps per 5 min. intervals and plot the average daily activity pattern.

````{r scatterplot2, fig.width=12}
agg2 <- with(data, aggregate(data$steps, by=list(data$interval), mean, na.rm = TRUE))
plot(agg2$Group.1, agg2$x, type = "l", xlab = "5 Minutes Intervals", ylab = "Steps Frequency", 
     main="Average daily activity pattern")
```

```{r max, echo=FALSE}
interval <- agg2$Group.1[which.max(agg2$x)]
average <- max(agg2$x)
```

The interval with the highest averaged number of steps is as follows: `r interval`. The highest averaged number of steps is `r round(average, digit=2)`, as shown in the graf above.

## Imputing missing values
The total number of "NA" (missing values) is:
```{r}
sum(is.na(data [,1]))
```
We replace missing values with the "5 minutes interval" mean and create a new dataset.
```{r}
names(agg2)[names(agg2) == 'Group.1'] <- 'interval'
names(agg2)[names(agg2) == 'x'] <- 'newsteps'
data2 <- merge(data, agg2, by = "interval")
data2$steps[is.na(data2$steps)] <- data2$newsteps[is.na(data2$steps)]
##Create a new dataset
newdata <- subset(data2, select = c("steps","interval","date2"))
```
Then, we re-aggregate the total number of steps taken per day and create a histogram that shows it.
````{r scatterplot3, fig.width=12}
agg3 <- with(newdata, aggregate(newdata$steps, by=list(newdata$date2), sum))
barplot(agg3$x, names.arg = agg3$Group.1, col="red", main="Total number of steps taken per day (with imputed value)", xlab="Days: From October 1st to December 1st",ylab="Step Frequency")
```
We calculate the mean and the median, as shown respectively:
```{r}
mean(agg3[, "x"], na.rm = F)
median(agg3[, "x"], na.rm = F)
```
Conclusion: As expected, the effect of replacing missing values with the "5 min. interval" mean has no impact on the mean but has a very low effect on the median, equaling it exactly to the mean.

## Are there differences in activity patterns between weekdays and weekends?
We create a new variable showing the type of week days.
```{r}
## Create a new variable (factor) for type of days
newdata$date <- weekdays(newdata$date2)

newdata$daytype <- c("Weekday")
for (i in 1:nrow(newdata)){
  if (newdata$date[i] == "Saturday" || newdata$date[i] == "Sunday")
          {
    newdata$daytype[i] <- "Weekend"
          }
}
newdata$daytype <- as.factor(newdata$daytype)

# Aggregate the total number of steps taken per day.
agg4 <- with(newdata, aggregate(newdata$steps, by=list(newdata$interval, newdata$daytype), mean))
```

# Make the panel plot for weekdays and weekends
Finally, we plot the average daily activity pattern for weekdays and weekends.
```{r qplot, fig.width=12}
library(ggplot2)
qplot(Group.1, x, data=agg4, geom=c("line"), xlab="5 Minutes Intervals", 
      ylab="Averaged Number of steps", main="Average daily activity pattern for Weekday and Weekend (with imputed values)") + facet_wrap(~ Group.2, ncol=1)
```
Conclusion: The averaged daily activity patterns seem to be more active (with a highter number of steps) for the weekends than the weekdays on average as confirmed by the means for weekday and weekend below:
```{r}
tapply(agg4$x, agg4$Group.2, mean, na.rm = F)
```
